---
title: "Ch1 data visualization with ggplot2"
output: html_document
date: "2023-04-24"
---


Import \> tidy \> transform, visualize, model \> communicate

-   ggplot2 \> implements the grammar of graphics, building graph

```{bash}
library(tidyverse)
```

**First step**
- do cars with big engines use more fuel than cars with small engines?
(what does the relationship between engine size and fuel efficiency look)

```{r}
ggplot2::mpg
```

- displ > car engine size in liters 
- hwy > car fuel efficiency on highway in mpg

You can add 
```{r}
?mpg
```

to open help page to get more info

**Creating pplot**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot above shows a negative relationship between the size(displ)and fuel efficiency(hwy)
> therefore, cars with big engines use more fuel 

*how to begin a ggplot2*
- ggplot() will create a coordinate system to add layers too 
- ggplot(data=) will create an empty plot, also the first argument for the plot 
- geom_point() is a function that adds a layer of points to the plot, this is a scatterplot
- geom will take mappings argument that will define how variables are mapped to visual properties 
- geom is always paired with aes() that tells which variables to map to x and y axis

**graphing template**

```{r}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

**exercise**

plot hwy against cly 
```{r}
ggplot(data=mpg)+ geom_point(mapping = aes(x=hwy, y=cyl))
```

- use ?mpg to find out what variable means what

**aesthetic mapping**

https://learning.oreilly.com/api/v2/epubs/urn:orm:book:9781491910382/files/assets/rfds_01in02.png
- one group of points are outliers 
- we can test if the cars are hybrids. look at class value for each car. we can add a third value by mapping to an aesthetic

- an *aesthetic* is a visual property that includes, size, shape, or color 

- we can use level to describe aesthetic properties 

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

```

- we can map the colors of the class variables 
- *aesthetic to aes()* 
- ggplot2 will assign a level of aesthetic to each value > scaling 

- we can map class to size as well
- or we can map class to *alpha* aesthetic that controls transparency of the points

```{r}
Top
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

# Bottom
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

- ggplot2 only use 6 shapes at a time so any additional groups will go unplotted using this aesthetic

- you can set aesthetic properties of geom, ie, making all points blue 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
- this will go outside of aes()

**exercise**

1) what is wrong with this code and why are the points not blue?
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

- the ) should go after y input and not after the color input so it should be 
```{r}
ggplot(data=mpg)+geom_point(mapping=aes(x=displ, y=hwy), color= "blue")
```

- color= blue should not be inlcuded within the mapping argument


2) which variable in mpg are categorical which are continuous?

- type ?mpg
- those with <chr> are categorical and those with <dbl> or <int> are continuous 

3) map a continuous variable to color, size and shape how do they differ? 

-continuous variable when mapped using color : will scale from light to dark color that varies 
-size of the points will depends on function of their size 
-shape will give an error, numbers have orders but shapes do not 
4) what happens when you map the same variable to many different aes?

- it is reduent most of the time so we usually will avoid it 

5) what does stoke aes do? what shapes does it work with

- stroke changes thr size of the border for shapes 


**common problems**

-if left hand side of the console gives + then R doesnt think you typed a complete expression
- when creating ggplot2 make sure the + comes at the end and dont start a line with it 

**facets**

-you can split plots into facets, subplots for each display 
- use facet_wrap() to plot by single variable
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_wrap(~class, nrow=2)
```

-use ~ followed by variable names 

-to facet plot on two varibles add facet_grid() to your plot, it is also a formula that will contain two variable names separated by ~
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_grid(drv~cyl)
```

**exercise**

1) when you facet continuous var it will be converted to categorical var and the plot contains a facet for each value

2) empty plots in facet means that there are no observations there 

3) by using . it ignores the dimension when faceting. so it can ignore x or y input 

4) nrow and ncol are used when using facet_wrap() to layout the facets but its not nesscarry when it is for grid
-use ?facet_wrap to see more 

**geometric objects**
-if plots used different visual object then we say they are different geoms 
-geoms; geometrical object that a plot uses to represent data 
-every geom takes a mapping argument but not every aes works with every geom 
-there are extensions and cheatsheets