---
title: "Ch3 data visualization with ggplot2"
output: html_document
date: "2023-04-24"
---


Import \> tidy \> transform, visualize, model \> communicate

-   ggplot2 \> implements the grammar of graphics, building graph

```{bash}
library(tidyverse)
```

**First step**
- do cars with big engines use more fuel than cars with small engines?
(what does the relationship between engine size and fuel efficiency look)

```{r}
ggplot2::mpg
```

- displ > car engine size in liters 
- hwy > car fuel efficiency on highway in mpg

You can add 
```{r}
?mpg
```

to open help page to get more info

**Creating pplot**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot above shows a negative relationship between the size(displ)and fuel efficiency(hwy)
> therefore, cars with big engines use more fuel 

*how to begin a ggplot2*
- ggplot() will create a coordinate system to add layers too 
- ggplot(data=) will create an empty plot, also the first argument for the plot 
- geom_point() is a function that adds a layer of points to the plot, this is a scatterplot
- geom will take mappings argument that will define how variables are mapped to visual properties 
- geom is always paired with aes() that tells which variables to map to x and y axis

**graphing template**

```{r}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

**exercise**

plot hwy against cly 
```{r}
ggplot(data=mpg)+ geom_point(mapping = aes(x=hwy, y=cyl))
```

- use ?mpg to find out what variable means what

**aesthetic mapping**

https://learning.oreilly.com/api/v2/epubs/urn:orm:book:9781491910382/files/assets/rfds_01in02.png
- one group of points are outliers 
- we can test if the cars are hybrids. look at class value for each car. we can add a third value by mapping to an aesthetic

- an *aesthetic* is a visual property that includes, size, shape, or color 

- we can use level to describe aesthetic properties 

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

```

- we can map the colors of the class variables 
- *aesthetic to aes()* 
- ggplot2 will assign a level of aesthetic to each value > scaling 

- we can map class to size as well
- or we can map class to *alpha* aesthetic that controls transparency of the points

```{r}
Top
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

# Bottom
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

- ggplot2 only use 6 shapes at a time so any additional groups will go unplotted using this aesthetic

- you can set aesthetic properties of geom, ie, making all points blue 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
- this will go outside of aes()

**exercise**

1) what is wrong with this code and why are the points not blue?
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

- the ) should go after y input and not after the color input so it should be 
```{r}
ggplot(data=mpg)+geom_point(mapping=aes(x=displ, y=hwy), color= "blue")
```

- color= blue should not be inlcuded within the mapping argument


2) which variable in mpg are categorical which are continuous?

- type ?mpg
- those with <chr> are categorical and those with <dbl> or <int> are continuous 

3) map a continuous variable to color, size and shape how do they differ? 

-continuous variable when mapped using color : will scale from light to dark color that varies 
-size of the points will depends on function of their size 
-shape will give an error, numbers have orders but shapes do not 
4) what happens when you map the same variable to many different aes?

- it is reduent most of the time so we usually will avoid it 

5) what does stoke aes do? what shapes does it work with

- stroke changes thr size of the border for shapes 


**common problems**

-if left hand side of the console gives + then R doesnt think you typed a complete expression
- when creating ggplot2 make sure the + comes at the end and dont start a line with it 

**facets**

-you can split plots into facets, subplots for each display 
- use facet_wrap() to plot by single variable
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_wrap(~class, nrow=2)
```

-use ~ followed by variable names 

-to facet plot on two varibles add facet_grid() to your plot, it is also a formula that will contain two variable names separated by ~
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_grid(drv~cyl)
```

**exercise**

1) when you facet continuous var it will be converted to categorical var and the plot contains a facet for each value

2) empty plots in facet means that there are no observations there 

3) by using . it ignores the dimension when faceting. so it can ignore x or y input 

4) nrow and ncol are used when using facet_wrap() to layout the facets but its not nesscarry when it is for grid
-use ?facet_wrap to see more 

**geometric objects**
-if plots used different visual object then we say they are different geoms 
-geoms; geometrical object that a plot uses to represent data 
-every geom takes a mapping argument but not every aes works with every geom 
-there are extensions and cheatsheets
- geom_smooth uses a single geometric object to display rows of data 
-you can add multiple geoms to the same plot 
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  geom_smooth(mapping=aes(x=displ, y=hwy))
```
-but this will introduce some duplication to the code but we can fix that by passing a set of mapping to ggplot 
```{r}
ggplot(data=mpg,mapping=aes(x=displ, y=hwy))+
  geom_point()+
  geom_smooth()
```
- you can display different aes in different layers 
```{r}
ggplot(data=mpg, mapping=aes(x=displ, y=hwy))+
  geom_point(mapping=aes(color=class))+
  geom_smooth()
```
**exercies**
1)line chart: geom_line 
boxplot: geom_boxplot()
histogram: geom_histogram()
area chart: geom_area()

2) show.legend=FALSE will hide the legend box. you can remove it or show.legend=TRUE to display it

4) se added to arguments like geom_smooth will add standard error bands to the lines. by default without se=TRUE it will still be displayed 

**statistical transformations**

-geom_bar for bar chart
```{r}
ggplot(data=diamonds)+
  stat_count(mapping=aes(x=cut))
```
-every geom has a default stat and every stat has a default geom
-you can typically use geoms without statisical transformations 
-you might need to use stat when:

1)you want to override the default stat 
2)you want to override default mapping from transformed variables to aes, ie, display bar chart of proportion rather than count 
3) you want to draw greater attention to stat transformation in code
-stat_(summary) will summarize the y value for each x value
```{r}
ggplot(data=diamonds)+
  stat_summary(
    mapping=aes(x=cut, y=depth),
    fun.min=min,
    fun.max=max,
    fun=median
  )
```
= ?stat_bin got list of stats 

**exercise**
1)the default geom for stat_summary() is geom_pointrange(). the geom_pointrange() default stat is identity()
- we can add stat="summary" to use stat_summary() instead of stat_identity()
```{r}
ggplot(data=diamonds)+
  geom_pointrange(
    mapping=aes(x=cut, y=depth),
    stat="summary"
  )
```
-we need to specify values for fun.min, fun.max and fun
```{r}
ggplot(data=diamonds)+
  geom_pointrange(
    mapping=aes(x= cut, y=depth),
    stat="summary",
    fun.min=min,
    fun.max=max,
    fun=median
    
  )
```
2)geom_col() has a different default stat than geom_bar() and the default is stat_identity() that will leave the data as is.
geom_bar() default stat is stat_count() and only expects an x variable 
4)stat_smooth()
y= predicted value
ymin=lower value of the confidence interval
ymax= higher value of the confidence interval
se= standard error 
5) if group=a is not included then all the bars in the plot will have the same height. geom_bar() assumes that groups are equal to x 

**3.8 position adjustments**
- you can color a bar chart using color aes or fill 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
- mapping fill aes to antoher variable like clarity the bars are stacked and each color represents a combo of cut and clarity
- stacking is performed by position adjustment specified by the position argument
-if you dont want to stack bar chart then use identity, dodge or fill 
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```
-identity will place each object exactly where it falls in context of the graph
-alpha will make the graph transparent if you use fill=NA it will make it completely transparent 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```
- fill for position works like stacking but makes each set of stacked bars the same height
-easier to compare proportions across groups 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```
-dodge will place overlapping objects beside one another that makes it easier to compare individual values 

- overplotting can be an issue where not all the points show on graph 
-can be avoided position="jitter" that will add small noise to each point that will spread the point out 
-makes it less accurate at small scales but more revealing for large scales 
- geom_jitter()

**exercise**
2) width and height arguments for jitter parameters 
3) geom_jitter adds random variation to the locations points of the graph that reduces overlapping since two points are unlikely to have the same random variation 
-geom_count sizes the points relative to the number of observations, it will not change the coordinates of the points but can create overplotting 
4)default position for geom_plot is "dodge2" that is a shortcut for "position_dodge2"
-moves the geom horizontally to avoid overlapping other geoms 

**3.9 coordinate systems**
-the default is cartesian 
-coord_flip() switches x and y useful if you want horizontal boxes
-coord_quickmap() sets the aspect ratio correctly for maps, important if plotting spatial data 
-coord_polar() uses polar coord that reveal connection between bar chart and coxcomb chart

**excercies**
1)turn a stacked bar chart into a pie chart using coord__polar()
- pie chart is a stacked bar chart with polar coord
```{r}
ggplot(mpg, aes(x=factor(1), fill=drv))+
  geom_bar()
```
```{r}
ggplot(mpg, aes(x=factor(1),fill=drv))+
  geom_bar(width=1)+
  coord_polar(theta="y")
```
-adding theta="y" will create the pie chart taht maps y to the angle of each section
```{r}
ggplot(mpg, aes(x=factor(1),fill=drv))+
  geom_bar(width=1)+
  coord_polar()
```
-without theta it will create a bullseye chart 
2)labs() function adds axis titles, plot titles, and a caption.the argument are optional so you can add as many or few
3) coord_map() uses map projections to project 3d into 2d, mercator projection that applies to all geoms in the plot
- coord_quickmap() project faster, coord of geoms dont need to be transformed 
4) coord_fixed() ensures that the line produced by geom_abline() is at a 45 degree angle 

**3.10 layered grammar of graphics**

```{r}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

