---
title: "Ch1 data visualization with ggplot2"
output: html_document
date: "2023-04-24"
---


Import \> tidy \> transform, visualize, model \> communicate

-   ggplot2 \> implements the grammar of graphics, building graph

```{bash}
library(tidyverse)
```

**First step**
- do cars with big engines use more fuel than cars with small engines?
(what does the relationship between engine size and fuel efficiency look)

```{r}
ggplot2::mpg
```

- displ > car engine size in liters 
- hwy > car fuel efficiency on highway in mpg

You can add 
```{r}
?mpg
```

to open help page to get more info

**Creating pplot**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot above shows a negative relationship between the size(displ)and fuel efficiency(hwy)
> therefore, cars with big engines use more fuel 

*how to begin a ggplot2*
- ggplot() will create a coordinate system to add layers too 
- ggplot(data=) will create an empty plot, also the first argument for the plot 
- geom_point() is a function that adds a layer of points to the plot, this is a scatterplot
- geom will take mappings argument that will define how variables are mapped to visual properties 
- geom is always paired with aes() that tells which variables to map to x and y axis

**graphing template**

```{r}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

**exercise**

plot hwy against cly 
```{r}
ggplot(data=mpg)+ geom_point(mapping = aes(x=hwy, y=cyl))
```

- use ?mpg to find out what variable means what

**aesthetic mapping**

