---
title: "ch5 data transformation"
output: html_document
date: "2023-06-19"
---

**5.1 intro**
- focus on how to use dplyr package 
- the flights data prints different because of tibble
- tibbles are data frames but tweaked to work better in tidyverse 
- varibles types : int, dbl, chr, dtt, lgl, fctr, date
```{r}
flights
```


*dplyr basics*
-pick observations by value filter()
-reorder the rows arrange()
-pick variables by name selec()
-create new variables with functions of existing variables mutate()
-collapse values down to single summary summaries()

1) first argument is a data frame 
2) subsequent arguemnts describes what to do with the data frames using the variable names 
3) result is a new data frame 

**5.2 filter rows**
- allows you to subset observations base on their values 
```{r}
filter(flights, month == 1, day == 1)
```

- first argument is the name of data frame
-second and subsequent arguments are expressions that filter the data frame 
- filter(flights, month == 1, day == 1) will select flights from 1/1
-dplyr will filter and return a new data set but they will never modify inputs
- if you want to save the result you need to use assignment operations 
```{r}
(jan1 <- filter(flights, month == 1, day == 1))
jan1
#
```
- two () will print and save the variable 
- need to know how to select observations you want using comparison operators 
- R provides the standard suite: >, >=, <, <=, != (not equal), and == (equal).
- use == and not = to test for equality
-you can use near() instead of == 

- filter are combined with and so every expression must be true in order for a row to be included in the output
- & is “and”, | is “or”, and ! is not

```{r}
filter(flights, month == 11 | month == 12)
```
```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```

- order of operations doesnt work like english 
- x %in% y will select every row where x is one of the values in y
- you can simplify complicated subsetting by dr morgan's law: !(x & y) is the same as !x | !y, and !(x | y) is the same as !x & !y
```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```
- finding flights that werent delayed by more than two hours 
- make complicated multipart expressions in filter() explicit variables instead that is easier to check work 

NA or missing values can make comparison tricky 
- NA represents unknown value so missing values are contagious
- almost any operation involving unknown value will be unknown 
```{r}
NA > 5
#> [1] NA
10 == NA
#> [1] NA
NA + 10
#> [1] NA
NA / 2
#> [1] NA
```
if x and y are unknowns then they cannot be the same therefore x == y isnt good 
- if you want to determine if a value is missing use is.na()
- filter() only inlcudes rows where condition is TRUE, it excludes false and Na

**exercises**


