---
title: "ch5 data transformation"
output: html_document
date: "2023-06-19"
---

**5.1 intro**
- focus on how to use dplyr package 
- the flights data prints different because of tibble
- tibbles are data frames but tweaked to work better in tidyverse 
- varibles types : int, dbl, chr, dtt, lgl, fctr, date
```{r}
flights
```


*dplyr basics*
-pick observations by value filter()
-reorder the rows arrange()
-pick variables by name selec()
-create new variables with functions of existing variables mutate()
-collapse values down to single summary summaries()

1) first argument is a data frame 
2) subsequent arguemnts describes what to do with the data frames using the variable names 
3) result is a new data frame 

**5.2 filter rows**
- allows you to subset observations base on their values 
```{r}
filter(flights, month == 1, day == 1)
```

- first argument is the name of data frame
-second and subsequent arguments are expressions that filter the data frame 
- filter(flights, month == 1, day == 1) will select flights from 1/1
-dplyr will filter and return a new data set but they will never modify inputs
- if you want to save the result you need to use assignment operations 
```{r}
(jan1 <- filter(flights, month == 1, day == 1))
jan1
#
```
- two () will print and save the variable 
- need to know how to select observations you want using comparison operators 
- R provides the standard suite: >, >=, <, <=, != (not equal), and == (equal).
- use == and not = to test for equality
-you can use near() instead of == 

- filter are combined with and so every expression must be true in order for a row to be included in the output
- & is “and”, | is “or”, and ! is not

```{r}
filter(flights, month == 11 | month == 12)
```
```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```

- order of operations doesnt work like english 
- x %in% y will select every row where x is one of the values in y
- you can simplify complicated subsetting by dr morgan's law: !(x & y) is the same as !x | !y, and !(x | y) is the same as !x & !y
```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```
- finding flights that werent delayed by more than two hours 
- make complicated multipart expressions in filter() explicit variables instead that is easier to check work 

NA or missing values can make comparison tricky 
- NA represents unknown value so missing values are contagious
- almost any operation involving unknown value will be unknown 
```{r}
NA > 5
#> [1] NA
10 == NA
#> [1] NA
NA + 10
#> [1] NA
NA / 2
#> [1] NA
```
if x and y are unknowns then they cannot be the same therefore x == y isnt good 
- if you want to determine if a value is missing use is.na()
- filter() only inlcudes rows where condition is TRUE, it excludes false and Na

**exercises**
1) had an arrival delay two or more hours 
use arr_delay 
```{r}
filter(flights, arr_delay >= 120)
```
the flights that flew to houston are those flights where the destination is either IAH or HOU
-using %in% would be more compact and would scale to other airports 
```{r}
filter(flights, dest=="IAH"|dest=="HOU")
```
```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

operated by united american or delta 
- the column carrier indicates airline but uses codes that can be found in airline dataset 
```{r}
airlines
```
```{r}
filter(flights, carrier %in% c("AA", "DL", "UA"))
```

departed in summer (july august september)
```{r}
filter(flights, month>=7, month<=9)
```
- %in% operator is an alternative if the : operator is used to specify integer range the expression is readable and compact
```{r}
filter(flights, month %in% 7:9)
```

arrived more than two hours late but didnt leave late 
```{r}
filter(flights, arr_delay>120, dep_delay<=0)
```

2) another useful dplyr filtering helper is between()
- can be used to simplify operators 

3)how many flights have a missing dep_time what other variables are missing 

4) NA|TRUE is true because anything is true is true 
is its false its false if its true its true 

**5.3 arrange rows with arrange()**
- works like filter but it changes the orders of the rows 
- takes a data frame and set column names to order by
```{r}
arrange(flights, year, month, day)
```

- use desc() to recorder by column in descending order 
```{r}
arrange(flights, desc(dep_delay))
```

- missing values are always sorted at the end 

**exercises**
1) how can you use arrange() to sort all missing values to the start
- arrange puts NA last and using desc() will not change that 
- we can add indicator of whether the column has missing value then sort by missing indicator column 
```{r}
arrange(flights, desc(is.na(dep_time)), dep_time)
```

2) sort flights to find most delayed flighs and find those that left earliest 

```{r}
arrange(flights, desc(dep_delay))
``` 
- HA was the most delayed flight 

```{r}
arrange(flights, dep_delay)
```
- B6 departed earliest

3) sort flights to find fastest flights 
```{r}
head(arrange(flights, air_time))
```

- head() is used to get the first rows of data and tail() for the last 

**5.4 select columns with select()**
- starts_with("abc"): matches names that begin with “abc”.

--ends_with("xyz"): matches names that end with “xyz”.

contains("ijk"): matches names that contain “ijk”.

-matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.

-num_range("x", 1:3): matches x1, x2 and x3.
- use rename() which is a variant of select 
```{r}
rename(flights, tail_num = tailnum)
```

- use everything() if you have handful of variables you want to move to start of the data 

**exercises**

2) what happens if you include the name of a variable many times in select
- ignores duplication so they are only included once 

3) one_of()?? 
- selects variables with character vector rather than unquoted variable name arguements 
- easier to generate character vectors with variable names 

**5.5 add new variables with mutate()**
- mutate() always adds new columns at the end of data set 
- easiest way to see all column is view()
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```
- you can refer to columns you created by 
```{r}
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
- use transmute() if you only want to keep the new variables 